---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Spring 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Date objects

## Data Types in R

Data types include, numeric, decimals, boolean = true/false


R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

- every column is a vector and if they are the same length you can combine them all into 1 data frame

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector with floating point numbers bc decimal
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector, when you put it in quotes it makes it a string and the vector a character vector
vector3

vector1[3] # to retrieve a single value in the vector

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # 5 rows and four coulmns
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) # byrow puts the numbers in order by row
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line, dimnames gives names to the rows and columns
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) #
matrix3

matrix1[4, ] #matrix notation [row, column] 4th row
matrix1[ , 3] # 3rd column
matrix1[c(12, 14)] #
matrix1[c(12:14)] #
matrix1[2:4, 1:3] # rows 2-4 and columns 1-3

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]] #has 4 items in the list can use $ to see items in the list (key)

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? numeric
e <- c("red", "white", "red", NA) # What type of vector? character
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? logical 
dataframe1 <- data.frame(d,e,f) # creating the data frame
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # labeling the columns

dataframe1[1:2,] # shows rows 1-2
dataframe1[c("ID","Passed")] # shows columns labeled ID and Passed
dataframe1$ID # shows the items in the ID column


```
Question: How do the different types of data appear in the Environment tab?

> Answer: 

Lists, matrices and data frames 


Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: 

It shows you the list of items in yur data from for you to choose from.


### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}

temp.highs <- c(58, 41, 41, 46, 35, 38, 44, 48, 52, 56)
temp.highs

temp.lows <- c(31, 20, 25, 28, 19, 23, 25, 29, 43, 49)
temp.lows
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r}
(f -32) * 5/9 = c

celsius <- function(f){
  c <- (f-32)*5/9
  return(round(c,2))}

celsius(temp.highs)
celsius(temp.lows)
```

Combine your four vectors into a data frame and add informative column names.

```{r}

tempdatframe <- data.frame(temp.highs, temp.lows, celsius(temp.highs), celsius(temp.lows))

tempdatframe

names(tempdatframe) <- c("F Highs","F Lows","C Highs", "C Lows"); View(tempdatframe)
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}

summary(tempdatframe)

```

### Date objects

Remember formatting of dates in R: 

%d  day as number (0-31)
%m  month (00-12, can be e.g., 01 or 1)
%y  2-digit year
%Y  4-digit year
%a  abbreviated weekday
%A  unabbreviated weekday
%b  abbreviated month
%B  unabbreviated month

```{r}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented)
today <- Sys.Date()
class(today)
format(today, format = "%B")
<<<<<<< HEAD
#format(today, format = "")
#format(today, format = "") can use this to play around w dates
=======
format(today, format = "%y")
format(today, format = "%m-%d-%Y")
>>>>>>> b3c7106116c07efb5453f2e1b6ba3c74dfab9fdd
#format(today, format = "")

```


### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][https://cran.r-project.org/web/packages/lubridate/lubridate.pdf].

```{r DateExercise1}
#install.packages("lubridate")
library(lubridate)

#Ex1
str_today <- "2021-feb-3"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today

#Ex2
str_today <- "21-feb-3"
#Sine the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today

#there are other similar functions ydm(), mdy(), etc
```

```{r DateExercise2}
#century issue
str_past <- "55-feb-3"
date_obj_past <- ymd(str_past)
date_obj_past

#Build a function to fix year that is more general than the one discussed in the lesson
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d)
}

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=21) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue
str_past <- "55-feb-3"
#Alternative 1
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=22L)
date_obj_past

#Alternative 2
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=22L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine. 

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin

```