---
title: "Assignment 3: Data Exploration"
author: "Sara Diamond, Section #2"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
getwd() # looking at WD
setwd("/Users/saradiamond/Documents/Environmental_Data_Analytics_2022") #setting correct WD
library(tidyverse) # loading packages
library(ggplot2) #loading packages
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE) 
#reading in data ecotox file
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
#loading in data for litter and woody debris
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: 

According to Cornell University's College of Agriculture and Life Sciences, Neonicotinoids are a commonly used insecticide and are very popular in agricultural practices. While they can be less toxic to vertebrates than other insecticides, there is some worry of the effects they might have on pollinators like bees, which are crucial to every ecosystem. It would be important to know the ecotoxicology in order to determine the impacts they might have on these insects and how we can lessen them.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: 

It is important to understand woody debris and litter because it is adds nutrients to the soil of a forest ecosystem as well as plays a role in moisture retention. It can also provide habitats for smaller terrestrial species.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* The litter and debris were collected from elevated ground traps
* sampling takes place where there woody vegetation >2m tall
* Ground traps were sampled one time per year



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r, echo = TRUE}

summary(Neonics)
#summary of the entire dataset
dim(Neonics) 
#finding the number of rows and columns
```
> Answer: 

There are 4623 rows and 30 columns in this dataset.

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect) 
#looking specifically at the effects column
```

> Answer: 

The most common effects are population, mortality, behavior, feeding behavior, and reproduction. These might be the most interesting to look at because some are major parts of life stages and if they are having negative effects then it also may be having impacts on other areas as well and on other animal and plant species besides pollinators. 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r, echo = TRUE}

summary(Neonics$Species.Common.Name) 
#summary of the species

```

> Answer: 

The six most commonly studied species are the Honey Bee, the Parasitic Wasp, the Buff Tailed Bumblebee, the Carniolan Honey Bee, the Bumble Bee and the Italian Honeybee. All 6 of these species are pollinators. As stated previously, neonicotinoids are of concern because of the major impacts they have specifically on pollinator species. Pollinators serve a crucial role in the production of plants and without them, it would be disastrous. It is important to study these species because then you could have an opportunity to make management decisions that would have positive impacts for pollinations.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r, echo = TRUE}

class(Neonics$Conc.1..Author.) 
#determining the class of the Conc.1..Author column in the dataset. 

```

> Answer:

The class of the Conc.1..Author column is a factor. It is not numeric because even thought the column contains number, some of the entries has slashes, symbols or letters, which makes R think that this is not a numeric class. If all the entries were numbers without any other symbols or letters then it would be classified as a numeric class.

## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r, echo = TRUE}

ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50) 
#creating a frequency plot of the number of publications per year


```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location)
                , bins = 50) #making the same graph but now denoting colors to specific test locations


```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 

Based on the graph, it seems the most common testing locations were in the lab, with a greenhouse and indoor pots, followed by field natural, which means the "test system and the stressor are naturally derived."

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() #creating bar graph of the enpoint column to show counts
```

> Answer: 

The two most common endpoint counts are NOEL and LOEL. NOEL is defined as no observable effect level, which means that the highest dose is not creating effects that are significantly different from the different responses of the controls. LOEL is defined as lowest observable effect level, which means the lowest dose that had effects were significantly different than that of the controls used. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}

class(Litter$collectDate) 
#checking the class of date column and it is a factor

Litter$collectDate <- as.Date(Litter$collectDate) 
#changing to class to date

class(Litter$collectDate) 
#checking to make sure it is now a date class

unique(Litter$collectDate)
#determining which dates in August 2018 were data collected - August 2, 2018 and August 30, 2018

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

unique(Litter$plotID) 
#seeing how many plots at Niwot Ridge

summary(Litter$plotID)
#using summary tool for comparison



```

> Answer: 

The unique tool shows the different categories within that one column whereas the summary tool shows the categories and the numbers of each. This shows that there were 12 different plots sampled at Niwot Ridge and eahc one was sampled a speicifc amount of times. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}


ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() 
#creating a bar graph of the types of litter collected at the sites 

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)) 
#box plot of the mass according to the litter types 

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75)) 
#creating violin plot of same data 

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: 

The boxplot shows the distribution of the continuous values of dry mass in this case while the violin plot shows the density of the values. In this case, the box plot might be better because we are simply just interested in some summary statistics of the data rather than the entire distribution of the data. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: 

Based on the plots, the needles and the mixed litter categories tend to have the highest biomass. This makes sense here because the Niwot Ridge LTER is located in the Rocky Mountains in a higher elevation forest, which consists of trees mostly with needles. 